### 2013-03-27 doc/sosp13/abstract.tex
edits to abstract and intro

\Grappa is a runtime system for commodity clusters of multicore computers that
presents a massively parallel, single address space abstraction to
applications. \Grappa's purpose is to enable scaling of irregular
parallel applications, such as graph processing. Poor data locality, imbalanced parallel
work and complex communication patterns make scaling these applications difficult.

\Grappa serves both as a C++ user library and as a foundation upon which
higher level languages can be developed or adapted. \Grappa tolerates delays
to remote memory by multiplexing thousands of lightweight workers
to each processor core; balances load via fine-grained distributed
work-stealing; increases communication throughput by aggregating
smaller data requests into large ones; and provides efficient synchronization
and remote operations. We present a detailed description of the \Grappa
system and
%programming examples using the library interface,
performance comparisons on several irregular benchmarks to hand-optimized MPI code and to
the Cray XMT, a custom system used to target the real time graph analytics
market.

>>>>

\Grappa is a runtime system for commodity clusters of multicore computers that
presents a massively parallel, single address space abstraction to
applications. \Grappa's purpose is to enable scalable performance of irregular
parallel applications, such as graph processing. Poor data locality,
imbalanced parallel work and complex communication patterns make scaling these
applications difficult.

\Grappa serves both as a C++ user library and as a foundation upon which
higher level languages can be developed or adapted. \Grappa tolerates delays
to remote memory by multiplexing thousands of lightweight workers to each
processor core; balances load via fine-grained distributed work-stealing;
increases communication throughput by aggregating smaller data requests into
large ones; and provides efficient synchronization and remote operations. We
present a detailed description of the \Grappa system and performance
comparisons on several irregular benchmarks to hand-optimized MPI code and to
the Cray XMT, a custom system used to target the real time graph analytics
market.

### 2013-03-27 doc/sosp13/intro.tex
edits to abstract and intro

Irregular applications exhibit workloads, dependences, and memory accesses
that are highly sensitive to input. Classic examples of such applications
include branch and bound optimization, SPICE circuit simulation, contact
algorithms in car crash analysis, and network flow. Important contemporary
examples include processing large graphs in the business, national security,
machine learning, and social network computing domains. For these emerging
applications, fast response -- given the sheer amount of data -- requires
multinode systems. The most broadly available multinode systems are those
built from x86 compute nodes interconnected via ethernet or InfiniBand.
Scalable performance of irregular applications on these systems is elusive
for two reasons:

\noindent{\bf Low Injection Rate} Reference patterns of irregular
applications are unpredictable, with odds increasing as data is
spread out more broadly across a larger system that a given reference
will refer to memory located remotely.  This behavior contrasts
sharply with that of data-parallel applications for which efficient
frameworks such as MapReduce~\cite{mapreduce:osdi04} may exploit the
structure of computation and data affinity to limit internode
communication to less infrequent, larger messages.  Mass
market network technology is designed to transfer these larger blocks
of data, not the word-sized references emitted by irregular
application tasks. Injection rate into the network is insufficient to
utilize wire bandwidth when blocks are below about two-kilobytes, so
any naive communication strategy severely under-utilizes the network.

\noindent{\bf High Latency} To tolerate latency, data parallel applications make explicit
their regular, predictable communications so as to minimize interference with computation.
In contrast, irregular application tasks encounter remote references dynamically during
execution and must resolve them before making further progress.  The latency of
commodity networks is anywhere from one to hundreds of microseconds,
far more than can be tolerated via the hyperthrading available in x86 processors.
Consequently, even with RDMA, processor prefetching and caching hardware alone is inadequate
to ensure pipeline utilization in the presence of high latency networks.

While some irregular applications can be restructured to better exploit
locality, aggregate requests to increase message size, and manage the
additional challenges of load balance and synchronization across multinode
systems, the effort required to do so is formidable and involves knowledge and
skills pertaining to distributed systems far beyond those of most application
programmers. Luckily, many of the important irregular applications (e.g.,
graph processing, our focus in this paper) naturally offer large amounts of
concurrency. This immediately suggests taking advantage of concurrency to
tolerate the latency of data movement.

The fully custom Tera MTA-2~\cite{tera:mta1} system is a classic example of
supporting irregular applications by using concurrency to hide latencies. It
had a large distributed shared memory with no caches. On every clock cycle,
each processor would execute a ready instruction chosen from one of its 128
hardware thread contexts, a sufficient number to fully hide memory access
latency. The network was designed with a single-word injection rate that
matched the processor clock frequency and sufficient bandwidth to sustain a
reference from every processor on every clock cycle. Unfortunately, the MTA
was not general enough nor cost-effective. The Cray XMT approximates the Tera
MTA-2, substantially reducing its cost but not overcoming its narrow range of
applicability.

>>>>

Irregular applications exhibit workloads, dependences, and memory accesses
that are highly sensitive to input. Classic examples of such applications
include branch and bound optimization, SPICE circuit simulation, and car crash
analysis. Important contemporary examples include processing large graphs in
the business, national security, and social network computing domains. For
these emerging applications, reasonable response time -- given the sheer
amount of data -- requires large multinode systems. The most broadly available
multinode systems are those built from x86 compute nodes interconnected via
ethernet or InfiniBand. However, scalable performance of irregular
applications on these systems is elusive for two reasons:

\noindent{\bf Poor data locality and frequent communication.} Data reference
patterns of irregular applications are unpredictable and tend to be spread
across the entire system. This results in frequent requests for remote data.
Caches are of little assistance because of low data re-use and spatial
locality. Prefetching is ineffective because request locations are not known
early enough. Data-parallel frameworks such as
MapReduce~\cite{mapreduce:osdi04} are also inappropriate because they rely on
being able to partition data and regular communication patterns. As a
consequence, irregular applications require frequent communication of
typically small pieces of data. In contrast, commodity networks are designed
for large packets and have just a fraction of their peak bandwidth when moving
small packets.

\noindent{\bf High network communication latency.} The performance challenges
of frequent communication is exacerbated by the network latency relative to
processor performance. Latency of commodity networks runs anywhere from one to
hundreds of microseconds, which is much more than superscalar execution with
sophisticated memory hierarchies can tolerate. Since irregular application
tasks encounter remote references dynamically during execution and must
resolve them before making further progress, stalls are frequent and lead
severely underutilized compute resources.

While some irregular applications can be restructured to better exploit
locality, aggregate requests to increase network message size, and manage the
additional challenges of load balance and synchronization across multinode
systems, the effort required to do so is formidable and involves knowledge and
skills pertaining to distributed systems far beyond those of most application
programmers. Luckily, many of the important irregular applications (e.g.,
graph processing, our focus in this paper) naturally offer large amounts of
concurrency. This immediately suggests taking advantage of concurrency to
tolerate the latency of data movement by overlapping computation with
communication.

The fully custom Tera MTA-2~\cite{tera:mta1} system is a classic example of
supporting irregular applications by using concurrency to hide latencies. It
had a large distributed shared memory with no caches. On every clock cycle,
each processor would execute a ready instruction chosen from one of its 128
hardware thread contexts, a sufficient number to fully hide memory access
latency. The network was designed with a single-word injection rate that
matched the processor clock frequency and sufficient bandwidth to sustain a
reference from every processor on every clock cycle. Unfortunately, the MTA
was not general enough nor cost-effective. The Cray XMT approximates the Tera
MTA-2, reducing its cost but not overcoming its narrow range of applicability.
